`gosset`: An R package for analysis and synthesis of ranking data in agricultural

### 2022-06-25 inst/paper/paper.Rmd
example1

---
title: '`gosset`: An R package for analysis and synthesis of ranking data in agricultural
  experimentation'
abstract: null
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    number_sections: no
    keep_tex: no
latex_engine: pdflatex
editor_options:
  markdown:
    wrap: 72
bibliography: paper.bib
csl: citationstyle.csl
header-includes:
- \usepackage{pdfpages}
- \usepackage{caption}
- \usepackage{xcolor}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{font=small}
---

@@

# Required Metadata {#required-metadata .unnumbered}

# Current code version {#current-code-version .unnumbered}

| **Nr.** | **Code metadata description**                                   | **Please fill in this column**                             |
|:-----------------|:--------------------------|:-------------------------|
| C1      | Current code version                                            | 0.4                                                    |
| C2      | Permanent link to code/repository used for this code version    | <https://github.com/AgrDataSci/gosset>                     |
| C3      | Code Ocean compute capsule                                      |                                                            |
| C4      | Legal Code License                                              | MIT                                                        |
| C5      | Code versioning system used                                     | Git                                                        |
| C6      | Software code languages, tools, and services used               | R                                                          |
| C7      | Compilation requirements, operating environments & dependencies |                                                            |
| C8      | If available Link to developer documentation/manual             | <https://agrdatasci.github.io/gosset/>                     |
| C9      | Support email for questions                                     | [kaue.desousa@inn.no](mailto:kaue.desousa@inn.no){.email} |

@@

Participatory experimentation approaches have been increasingly applied in agricultural research [@deRoo2019farm]. While collecting data in ranking format is uncommon in general
agricultural research settings, it is often collected in participatory experiments [@Coe2002AnalyzingRA]. Recently developed approaches for on-farm experimentation,
such as the tricot methodology are based on the collection of data in ranking format [@vanetten_beza_2019]. On the other hand, newly proposed approaches for synthesis of crop variety evaluation data largely depend on the analysis of ranking data [@brown2020data].
The analysis of ranking data requires the use of appropriate statistical models such as Plackett-Luce [@luce_individual_1959; @Plackett] or Bradley-Terry [@bradley1952rank]. Functionality for fitting Bradley-Terry and Plackett-Luce models are available in R with the packages BradleyTerry2 and PlackettLuce respectively [@BradleyTerry2; @Turner2020]. However, extended functionality was required for the entire data science workflow, which usually includes: (1) Data preparation and cleaning, (2) modelling and validation, and (3) results presentation. For (1) gosset provides functions for converting and preparing data into ranking or pairwise format required by the packages PlackettLuce and BradleyTerry2
respectively. For (2), gosset provides functions for model selection and validation using cross-validation. In the case of (3), enhanced functionality for plotting model results is provided by the gosset package.

@@

* `AIC` computes the Akaike Information Criterion @akaike1974] for a Bradley-Terry model fitted with BradleyTerry2 [@BradleyTerry2] or a Plackett-Luce model fitted with [@Turner2020].  
* `btpermute` deviance-based forward variable selection [@lysen2009permuted] procedure for Bradley-Terry models.
* `crossvalidation` performs k-fold cross-validation, where k could be specified by the user.The default is 10-fold. Folds can be provided as a vector for a custom cross-validation, such as blocked cross-validation.  
* `forward` executes forward variable selection with cross-validation.  
* `kendallTau` computes the Kendall-tau rank correlation coefficient between two rankings [@kendall_1938].  
* `kendallW` computes Kendall's W (coefficient of concordance) among observed rankings and those predicted by the Plackett-Luce model [@Kendall_1939].  
* `pseudoR2` computes goodness-of-fit metrics, such as McFadden's pseudo-R2 [@mcfadden1973conditional].

@@

To demonstrate the functionality of the gosset package, we use the `nicabean` dataset, which was generated with decentralized on-farm trials of common bean (*Phaseolus vulgaris* L.) varieties in Nicaragua over five seasons (between 2015 and 2016). Following the tricot  approach [@vanetten_beza_2019], farmers were asked to test in their farms three varieties of common bean. The varieties were randomly assigned as incomplete blocks of size three (out of 10 varieties). The farmers assessed which of the three varieties had the best and worst performance in eight traits (vigor, architecture, resistance to pests, resistance to diseases, tolerance to drought, yield, marketability, taste). The farmers also provided their overall appreciation about the varieties, i.e., which variety had the best and the worst performance based on the overall performance considering all the traits. To analyze the data, we use the Plackett-Luce model implemented in the R package *PlackettLuce* [@Turner2020].

The `nicabean` is a list with two data frames. The first, `trial`, contains the trial data with farmers’ evaluations, ranked from 1 to 3, with 1 being the higher ranked variety and 3 the lowest ranked variety for the given trait and incomplete block. The rankings in this dataset were previously transformed from tricot rankings (where participants indicate best and worst) to ordinal rankings using the function `rank_tricot()`. The second data frame, `covar`, contains the covariates associated to the on-farm trial plots and farmers. This example will require the packages PlackettLuce [@Turner2020], climatrends [@climatrends], and ggplot2 [@ggplot2].

``` {r, message = FALSE, eval = TRUE, echo = TRUE}
library("gosset")
library("PlackettLuce")
library("climatrends")
library("ggplot2")

@@

```{r, message = FALSE, eval = TRUE, echo = TRUE}
R <- vector(mode = "list", length = length(traits))

@@

```{r, message=FALSE, eval=TRUE, echo=TRUE}

@@

```{r, message=FALSE, eval=TRUE, echo=TRUE, fig.cap="Trait performance (log-worth) of bean varieties in Niragua. Variety 'Amadeus' is set as reference (log-worth = 0). Blue values indicate a superior performance of varieties for a given trait, compared to the reference.  Red values indicate a variety with weak performance for the given trait, compared to the reference."}
worth_map(mod[-baseline],
          labels = traits[-baseline], 
          ref = "Amadeus 77") +
  labs(x = "Variety",
       y = "Trait")
```

To consider the effect of climate factors on variety performance, we used climatic covariates to fit a 
Plackett-Luce tree. For brevity, we used the variables maximum night time temperature (maxNT) and  maximum day time temperature, but more variables can be used in a Plackett-Luce tree. Temperature data were obtained from the AgERA5 dataset, using the R package ag5Tools [@ag5tools]. The temperature data downloaded from AgERA5 were used for computing the climatic variables with the package *climatrends* [@climatrends]. In this case, we only use the rankings corresponding to the overall performance of varieties.

```{r , message=FALSE, eval=FALSE, echo=TRUE}
# temp <- temperature(covar[, c("longitude","latitude")], 
#                     day.one = covar[, "planting_date"],
#                     span = 80)
# 
# head(temp[, 1:2])
```

To be coupled with climatic covariates, the rankings should be made formatted as grouped rankings. For this
we used the function ``rank_numeric`` with the parameter ``group = TRUE``.

@@

```{r rel1, message=FALSE, eval=TRUE, echo=FALSE}
nicabean_rel <- reliability(pl, ref = "Amadeus 77")

kable(t(nicabean_rel), col.names = c("node 2", "node 3"), caption = "Reliability of evaluated varieties
      in each of the resulting nodes of the Plackett-Luce tree")

>>>>

---
title: '`gosset`: An R package for analysis and synthesis of ranking data in agricultural
  experimentation'
abstract: null
output:
#  word_document: default
  pdf_document:
    fig_caption: yes
    number_sections: no
    keep_tex: no
latex_engine: pdflatex
editor_options:
  markdown:
    wrap: 72
bibliography: paper.bib
csl: citationstyle.csl
header-includes:
- \usepackage{pdfpages}
- \usepackage{caption}
- \usepackage{xcolor}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{font=small}
---

@@

# Code metadata {#required-metadata .unnumbered}

| **Code metadata description**                 | **Please fill in this column**                             |
|:--------------------------|:--------------------------|
| Current code version                                            | 0.4                                                       |
| Code repository    | <https://github.com/AgrDataSci/gosset>                    |
| Legal code license                                              | MIT                                                       |
| Code versioning system used                                     | git                                                       |
| Software code languages, tools, and services used               | R                                                         |
| Compilation requirements, operating environments & dependencies | R                                                         |
| Link to developer documentation             | <https://agrdatasci.github.io/gosset/>                    |
| Support email for questions                                     | [kaue.desousa@inn.no](mailto:kaue.desousa@inn.no){.email} |

@@

Participatory experimentation approaches have been increasingly applied in agricultural research [@deRoo2019farm]. While collecting data in ranking format is uncommon in general agricultural research settings, it is often collected in participatory experiments [@Coe2002AnalyzingRA]. Recently developed approaches for on-farm experimentation, such as the tricot methodology are based on the collection of data in ranking format [@vanetten_beza_2019]. On the other hand, newly proposed approaches for synthesis of crop variety evaluation data largely depend on the analysis of ranking data [@brown2020data].

The analysis of ranking data requires the use of appropriate statistical models such as Plackett-Luce [@luce_individual_1959; @Plackett] or Bradley-Terry [@bradley1952rank]. Functionality for fitting Bradley-Terry and Plackett-Luce models are available in R with the packages BradleyTerry2 [@BradleyTerry2] and PlackettLuce [@Turner2020] respectively. However, extended functionality was required for the entire data science workflow, which usually includes: (1) Data preparation and cleaning, (2) modelling and validation, and (3) results presentation. For (1) gosset provides functions for converting and preparing data into ranking or pairwise format required by the packages PlackettLuce and BradleyTerry2 respectively. For (2), gosset provides functions for model selection and validation using cross-validation. In the case of (3), enhanced functionality for plotting model results is provided by the gosset package.

@@

* `AIC` computes the Akaike Information Criterion [@akaike1974] for a Bradley-Terry model [@BradleyTerry2] or a Plackett-Luce model [@Turner2020].  
* `btpermute` deviance-based forward variable selection [@lysen2009permuted] procedure for Bradley-Terry models.
* `crossvalidation` performs k-fold cross-validation, where k could be specified by the user.The default is 10-fold. Folds can be provided as a vector for a custom cross-validation, such as blocked cross-validation.  
* `forward` executes forward variable selection with cross-validation.  
* `kendallTau` computes the Kendall-tau rank correlation coefficient between two rankings [@kendall_1938].  
* `kendallW` computes Kendall's W (coefficient of concordance) among observed rankings and those predicted by the Plackett-Luce model [@Kendall_1939].  
* `pseudoR2` computes goodness-of-fit metrics, such as McFadden's pseudo-R2 [@mcfadden1973conditional].

@@

To demonstrate the functionality of the gosset package, we use the `nicabean` dataset, which was generated with decentralized on-farm trials of common bean (*Phaseolus vulgaris* L.) varieties in Nicaragua over five seasons (between 2015 and 2016). Following the tricot  approach [@vanetten_beza_2019], farmers were asked to test in their farms three varieties of common bean. The varieties were randomly assigned as incomplete blocks of size three (out of 10 varieties). The farmers assessed which of the three varieties had the best and worst performance in eight traits (vigor, architecture, resistance to pests, resistance to diseases, tolerance to drought, yield, marketability, and taste). The farmers also provided their overall appreciation about the varieties, i.e., which variety had the best and the worst performance based on the overall performance considering all the traits. To analyze the data, we use the Plackett-Luce model implemented in the R package *PlackettLuce* [@Turner2020].

The `nicabean` is a list with two data frames. The first, `trial`, contains the trial data with farmers’ evaluations, ranked from 1 to 3, with 1 being the higher ranked variety and 3 the lowest ranked variety for the given trait and incomplete block. The rankings in this dataset were previously transformed from tricot rankings (where participants indicate best and worst) to ordinal rankings using the function `rank_tricot()`. The second data frame, `covar`, contains the covariates associated to the on-farm trial plots and farmers. This example will require the packages PlackettLuce [@Turner2020], climatrends [@climatrends], chirps [@chirps] and ggplot2 [@ggplot2].

``` {r starting, message = FALSE, eval = TRUE, echo = TRUE}
library("gosset")
library("PlackettLuce")
library("climatrends")
library("chirps")
library("ggplot2")

@@

```{r rankings, message = FALSE, eval = TRUE, echo = TRUE}
R <- vector(mode = "list", length = length(traits))

@@

```{r PLmodel, message=FALSE, eval=TRUE, echo=TRUE}

@@

```{r worthmap, message=FALSE, eval=TRUE, echo=TRUE, fig.cap="Trait performance (log-worth) of bean varieties in Niragua. Variety 'Amadeus' is set as reference (log-worth = 0). Blue values indicate a superior performance of varieties for a given trait, compared to the reference.  Red values indicate a variety with weak performance for the given trait, compared to the reference."}
worth_map(mod[-baseline],
          labels = traits[-baseline], 
          ref = "Amadeus 77") +
  labs(x = "Variety",
       y = "Trait")
```

To consider the effect of climate factors on yield, we used climatic covariates to fit a Plackett-Luce tree. For simplicity, we used the total rainfall (Rtotal) derived from CHIRPS data [@Funk2015], and computed using the R packages chirps [@chirps] and [@climatrends]. However, additional covariates can be used in a Plackett-Luce tree, for example using temperature data from R package ag5Tools [@ag5tools] or nasapower [@nasapower].

We request the CHIRPS data using the package chirps. Data should be returned as a matrix. This process can take some minutes to be implemented.

```{r , message=FALSE, eval=FALSE, echo=TRUE}
dates <- c(min(covar[, "planting_date"]),
           max(covar[, "planting_date"]) + 70)

chirps <- get_chirps(covar[, c("longitude","latitude")], 
                     dates = as.character(dates),
                     as.matrix = TRUE,
                     server = "ClimateSERV")
```

Than we prepare the matrix to compute the rainfall indices using the function `rainfall()` from package climatrends.

To be linket to covariates, the rankings should be coerced to 'grouped rankings'. For this we used the function `group()` from PlackettLuce. We retain the ranking corresponding to yield.

@@

```{r rel1, message=FALSE, eval=TRUE, echo=FALSE}
# nicabean_rel <- reliability(pl, ref = "Amadeus 77")
# 
# kable(t(nicabean_rel), col.names = c("node 2", "node 3"), caption = "Reliability of evaluated varieties
#       in each of the resulting nodes of the Plackett-Luce tree")
